Informing policy via dynamic models: Cholera in Haiti

### 2024-02-08 edits.tex
Major edits to the introduction

>>>>

%They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes. 
\newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence. For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards. They used this review to develop general recommendations for policy-guiding modeling. This review is particularly relevant given that their recommendations are specifically intended for the modeling of neglected tropical diseases, which includes cholera. Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty. Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on effectively modeling dynamic systems.}

\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling. Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}. Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article. Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient. \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles. Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework. 
}

### 2024-02-08 ms.Rnw
Major edits to the introduction

This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
\old{We build on a multi-group modeling exercise by Lee et al. [1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak were available at various spatial scales, which resulted with 17 studies that considered spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community.}
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
\new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
\new{The data that were used in \cite{lee20}, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti. \TODO{How? This statement should be epanded or re-evaluated.}
Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.

@@

\new{\editMechModels}

@@

We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
\new{Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
This includes a comprehensive tutorial on performing inference with Model~3, a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
Finally, we conclude with a discussion of the results\new{, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.}

@@

@@

In this article we \old{calibrate} \new{follow \cite{lee20} by calibrating} the parameters of each of our models \old{by maximizing the likelihood} \new{using maximum likelihood}, as described in Eq.~\myeqref{eq:likedef}.
The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
\eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
Alternatively, some investigations obtained parameters from previous sources without calibrating to data.}

@@

\new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
\eic{I'M NOT SURE THIS IS A CONCLUSION. LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER. MOST OTHER PAPERS STUDY ONE MODEL. WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).
We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}

>>>>

This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
\old{We build on a multi-group modeling exercise by Lee et al. [1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community}
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
\new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
\new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
\new{The possibility of controling the Haiti cholera outbreak via OCV was considered by various research groups.} \TODO{Citations}. 
In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}

@@

@@

\new{\editModelingGuidanceFirst}

\new{\editModelingGuidanceSecond}

\new{Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.}
We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
\new{Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
This includes a comprehensive tutorial on performing inference with Model~3, a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
Finally, we conclude with a discussion of the results\new{, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.}

@@

\jwc{This doesn't quite feel like the right place for this either, but I can't seem to pin down exactly where that would be yet. Perhaps the discussion?}
\new{\editMechModels}

@@

In this article we \old{calibrate} \new{follow \cite{lee20} by calibrating} the parameters of each of our models \old{by maximizing the likelihood} \new{using maximum likelihood}, as described in Eq.~\myeqref{eq:likedef}.
The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
\eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.}

@@

% \new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
% \eic{I'M NOT SURE THIS IS A CONCLUSION. LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER. MOST OTHER PAPERS STUDY ONE MODEL. WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
% In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).
% We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
% However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
% These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
% Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
% This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}
